The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (8th edition -- distributed with CrypTool version 1.4.10)} \vspace{60pt}

### 2007-06-05 script/de/chapters/primes.tex
Skript: updates, C307 Engl, mutual links)

\hline
\end{tabular}
\caption{Die 20 größten Primzahlen und ihr jeweiliger Zahlentyp
         (Stand Mai 2007)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Kown-Primes}
\end{center}
\end{table} 
\footnotetext{\index{Zahlen!Sierpinski}\index{Seventeen or Bust SoB}%
Diese Zahl wurde am 26.3.2007 im verteilt rechnenden Internet-Projekt
"`Seventeen or Bust"' (SoB) (\href{http://www.seventeenorbust.com}{\texttt{http://www.seventeenorbust.com}})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
gefunden. Anders als das bekannte \hyperlink{GIMPS-project}{GIMPS-Projekt} (Kapitel~\ref{zahlentyp_mersenne}), das immer größere der unendlich vielen Primzahlen aufspürt, könnte SoB aber irgendwann mal die gesetzte Aufgabe vollständig erledigt haben.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsubsection{Mersennezahlen $f(n) = 2^n - 1$ für $ n $ prim}
    \index{Primzahl!Mersenne} \index{Mersenne!Mersenne-Primzahl}
    Wie \hyperlink{MersenneNumbers01}{oben} gesehen, liefert diese 
    Formel wohl relativ viele große Primzahlen, aber es kommt -- wie 
    für $n=11$ [$f(n)=2.047$] -- immer wieder vor, dass das Ergebnis 
    auch bei primen Exponenten \hyperlink{Mer-nums-not-always-prim}{nicht}
    prim ist. \\
    Heute kennt man alle Mersenne-Primzahlen mit bis zu ca. 4.000.000 
    Dezimalstellen (\hyperlink{M-39}{M-39}
    \index{Mersenne!Mersenne-Primzahl!M-39}): \\
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}     % be_2005_UPDATEN_if-new-mersenne-prime-appears

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection
    [Verallgemeinerte Mersennezahlen $f(k,n) = k \cdot 2^n \pm 1$]
    {Verallgemeinerte Mersennezahlen 
    $f(k,n) = k \cdot 2^n \pm 1 $ für $ n $ prim und $ k $ kleine Primzahlen}
% \quad für
    \label{generalized-mersenne-no1}
    Für diese 1. Verallgemeinerung der 
    Mersennezahlen\index{Mersenne!Mersennezahl!verallgemeinerte} 
    gibt es (für kleine $k$) 
    ebenfalls sehr schnelle Primzahltests (vgl. \cite{Knuth1981}).
    Praktisch ausführen lässt sich das zum Beispiel mit der Software 
    Proths von Yves Gallot\index{Gallot, Yves} 
    ({\href{http://www.prothsearch.net/index.html}
           {\tt http://www.prothsearch.net/index.html}}).

@@

    Details hierzu finden sich unter:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.cerias.purdue.edu/homes/ssw/cun}
               {\tt http://www.cerias.purdue.edu/homes/ssw/cun}
\end{itemize}

>>>>

\hline
\end{tabular}
\caption{Die 20 größten Primzahlen und ihr jeweiliger Zahlentyp
         (Stand Mai 2007)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\footnotetext{\index{Zahlen!Sierpinski}\index{Seventeen or Bust SoB}%
Diese Zahl wurde am 26.3.2007 im verteilt rechnenden Internet-Projekt
"`Seventeen or Bust"' (SoB) (\href{http://www.seventeenorbust.com}{\texttt{http://www.seventeenorbust.com}})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
gefunden. Anders als das bekannte \hyperlink{GIMPS-project}{GIMPS-Projekt} (Kapitel~\ref{zahlentyp_mersenne}), das immer größere der unendlich vielen Primzahlen aufspürt, könnte SoB aber irgendwann mal die gesetzte Aufgabe vollständig erledigt haben.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsubsection{Mersennezahlen $f(n) = 2^n - 1$ für $ n $ prim}
    \index{Primzahl!Mersenne} \index{Mersenne!Mersenne-Primzahl}
    Wie \hyperlink{MersenneNumbers01}{oben} gesehen, liefert diese 
    Formel wohl relativ viele große Primzahlen, aber es kommt -- wie 
    für $n=11$ [$f(n)=2.047$] -- immer wieder vor, dass das Ergebnis 
    auch bei primen Exponenten \hyperlink{Mer-nums-not-always-prim}{nicht}
    prim ist. \\
    Heute kennt man alle Mersenne-Primzahlen mit bis zu ca. 4.000.000 
    Dezimalstellen (\hyperlink{M-39}{M-39}
    \index{Mersenne!Mersenne-Primzahl!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}     % be_2005_UPDATEN_if-new-mersenne-prime-appears

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection
    [Verallgemeinerte Mersennezahlen $f(k,n) = k \cdot 2^n \pm 1$ / Proth-Zahlen]
    {Verallgemeinerte Mersennezahlen 
    $f(k,n) = k \cdot 2^n \pm 1 $ für $ n $ prim und $ k $ kleine Primzahl
    / Proth-Zahlen\footnotemark}
    \footnotetext{%
       Diese wurden nach dem französischen Landwirt François Proth (1852-1879)
       benannt. Berühmter noch als die Proth-Primzahlen dürfte das eng damit
       zusammenhängende Sierpinski-Problem\index{Zahlen!Sierpinski} sein,
       nämlich Zahlen k zu finden, so dass $k*2^n+1$ zusammengesetzt ist für
       alle $n>0$. Siehe Tab. \ref{L_n_Largest_Known-Primes}.
    }
    \label{generalized-mersenne-no1}

    Diese 1. Verallgemeinerung der 
    Mersennezahlen\index{Mersenne!Mersennezahl!verallgemeinerte} 
    erzeugt die sogenannten Proth-Zahlen\index{Zahlen!Proth-Zahl}.
    Dafür gibt es (für kleine $k$) 
    ebenfalls sehr schnelle Primzahltests (vgl. \cite{Knuth1981}).\\
    Praktisch ausführen lässt sich das zum Beispiel mit der Software 
    Proths von Yves Gallot\index{Gallot, Yves}:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
            {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

    Details hierzu finden sich unter:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.cerias.purdue.edu/homes/ssw/cun}
            {\tt http://www.cerias.purdue.edu/homes/ssw/cun}}
\end{itemize}

### 2007-06-05 script/en/chapters/crypto2020.tex
Skript: updates, C307 Engl, mutual links)

In 2005, that is 27 years after the invention of
RSA, Bahr, Boehm, Franke, and Kleinjung from
Bochum University managed to factor a 200 digit
RSA challenge number
(\url{www.mat.uniroma2.it/~eal/rsa640.txt}).  A
key with size originally thought to be secure for
a very long time was broken with a computation
that took them just five months.  This illustrates
the tremendous progress factoring technology has
made within the last 30 years.  This progress is
based on break-through mathematical ideas --- e.g.\
the number field sieve proposed by John
Pollard --- as well as significant developments in
computer hardware and software implementation
technology.

In 2000, Lenstra and Verheul developed an
extrapolation formula that is supposed to help us
forecast the security\index{Security!forecast}
one can achieve with RSA and
other important cryptographic schemes in the long
term (\url{www.keylength.com}).  The formula
suggests the use of 850 digit RSA moduli if one
wishes to protect data for the next 30 years.
This corresponds to a 3072 bit RSA key.

Yet, even a well thought out extrapolation formula
is no security guarantee!  At any time, a
brilliant mathematical idea can allow us to factor
large numbers easily, and destroy the security of
RSA.  In 1996, Peter Shor showed that a quantum
computer --- a new type of computer that leverages
the laws of quantum mechanics to speed up certain
types of computations --- can in principle be used
for the fast factorization of large numbers \cite{shor:1997}.
Despite intensive research in the area, it is
still too early to judge whether we are ever going
to be able to build quantum computers\index{Quantum computer} of
sufficient capacity to apply Shor's algorithm to
numbers of relevant size.  Recent announcements of
significant progress in this area made by the
start-up company D-Wave (\url{www.dwavesys.com})
have been greeted with a lot of scepticism, even
ridicule.

>>>>

In 2005, that is 27 years after the invention of
RSA, Bahr, Boehm, Franke, and Kleinjung from
Bochum University managed to factor a 200 digit
RSA challenge number
(\url{www.mat.uniroma2.it/~eal/rsa640.txt}).  A
key with size originally thought to be secure for
a very long time was broken with a computation
that took them just five months.  This illustrates
the tremendous progress factoring technology has
made within the last 30 years.  This progress is
based on break-through mathematical ideas --- e.g.\
the number field sieve proposed by John
Pollard --- as well as significant developments in
computer hardware and software implementation
technology.\footnote{%
Please compare chapter \ref{SecurityRSA}
\hyperlink{SecurityRSA}{Considerations regarding the
security of the RSA algorithm}, and especailly chapters
\ref{NoteFactorisation} and \ref{FactorisationResearch}.
}

In 2000, Lenstra and Verheul\index{Lenstra/Verheul}
developed an extrapolation formula that is supposed
to help us forecast the security\index{Security!forecast}
one can achieve with RSA and other important
cryptographic schemes in the long
term (\url{www.keylength.com}).  The formula
suggests the use of 850 digit RSA moduli if one
wishes to protect data for the next 30 years.
This corresponds to a 3072 bit RSA key.

Yet, even a well thought out extrapolation formula
is no security guarantee!  At any time, a
brilliant mathematical idea can allow us to factor
large numbers easily, and destroy the security of
RSA.  In 1996, Peter Shor showed that a quantum
computer --- a new type of computer that leverages
the laws of quantum mechanics to speed up certain
types of computations --- can in principle be used
for the fast factorization of large numbers \cite{shor:1997}.
Despite intensive research in the area, it is
still too early to judge whether we are ever going
to be able to build quantum computers\index{Quantum computer} of
sufficient capacity to apply Shor's algorithm to
numbers of relevant size.\footnote{%
Required qbits for attacks on RSA, DSA and ECDSA using key with a bit length n: \\
\vskip +1 pt
\begin{tabular}{|c|l|}
\hline
   RSA		&  2n + 3 \\
   DSA		&  2n + 3 \\
   ECDSA $2^n$	&  \~{}2n + 8 log n \\
   ECDSA p	&  \~{}6n \\
\hline
\end{tabular}
\vskip +6 pt
Please compare chapter 5.3 in
``SicAri -- Eine Sicherheitsplattform und deren Werkzeuge
für die ubiquitäre Internetnutzung, KB2.1 -- Abschlussbericht,
Übersicht über Angriffe auf relevante kryptographische Verfahren'',
version 1.0, Mai 17, 2005,
Prof. Dr. Johannes Buchmann et al., TUD-KryptC and cv cryptovision GmbH
(\href{http://www.cdc.informatik.tu-darmstadt.de/~schepers/kb\_21\_angriffe.pdf}
 {\tt http://www.cdc.informatik.tu-darmstadt.de/\~{}schepers/kb\_21\_angriffe.pdf}).
}
Recent announcements of
significant progress in this area made by the
start-up company D-Wave (\url{www.dwavesys.com})
have been greeted with a lot of scepticism, even
ridicule.

### 2007-06-05 script/en/chapters/introduction.tex
Skript: Updates accord. to Vollmer's feedback about authors; C307 added (in German only yet)
Skript: updates, C307 Engl, mutual links)

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
Again this edition of the script amended and updated some topics:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers
        (Mersenne and Fermat primes, ``M44'') \\ 
	(chap. \ref{spezialzahlentypen}, \ref{RecordPrimes};
         table \ref{L_n_Largest_Kown-Primes}, 
         page \pageref{L_n_Largest_Kown-Primes}), 
  \item the factorisation of big numbers (``C307'')\index{C307} 
        (chap. \ref{RSA-200}),
  \item the current record (April 2007) calculating the discrete logarithm 
        (the used sample power consisted of the first 160 digits of Euler's
        number e and the calculation was done modulo a 160 digit prime)
        (chap. \ref{L_Discrete_Logarithm}),
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}
Newly added is chapter \ref{Chapter_Crypto2020} by Prof. Buchman et al.:
{\bf Krypto2020}.

@@

At this point I'd like to thank explicitly seven people who particularly 
contributed to CrypTool\index{CrypTool}. Without their special talents
 and engagement CrypTool would not be what it is today:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Prof. Dr. Claudia Eckert,  Mr.\ Thomas Buntrock and Mr.\ Thorsten Clausius.
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

@@

Frankfurt (Germany), June 2006

>>>>

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
Again this edition of the script amended and updated some topics, e.g.:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers
        (Mersenne and Fermat primes, ``M44'' or the biggest non-Mersenne prime
         found in March 2007 when investigating Sierpinski numbers) \\ 
	(chap. \ref{spezialzahlentypen}, \ref{RecordPrimes};
         table \ref{L_n_Largest_Known-Primes}, 
         page \pageref{L_n_Largest_Known-Primes}), 
  \item the factorisation of big numbers like ``C307''\index{C307} 
        (chap. \ref{C307}),
  \item the current record (April 2007) calculating the discrete logarithm 
        (The used sample power consisted of the first 160 digits of Euler's
        number e and the calculation was done modulo a 160 digit prime)
        (chap. \ref{L_Discrete_Logarithm}),
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}
Newly added is chapter \ref{Chapter_Crypto2020} by Prof. Buchman et al.:
{\bf Krypto2020}.

@@

At this point I'd like to thank explicitly the following six people who
particularly contributed to CrypTool\index{CrypTool}. Without their special
talents and engagement CrypTool would not be what it is today:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Prof. Dr. Claudia Eckert and  Mr.\ Thomas Buntrock.
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

@@

### 2007-06-05 script/en/chapters/moderncryptography.tex
Skript: updates, C307 Engl, mutual links)

%--------------------------------------------------------------------
\hypertarget{RSAVerfahren}{}
\subsubsection[The RSA procedure]
{The RSA procedure\footnotemark}
\footnotetext{
Using CrypTool\index{CrypTool} you can gain practical experience
with the RSA procedure via the menu {\bf Indiv.Procedures \textbackslash{} 
RSA Cryptosystem \textbackslash{} RSA Demonstration}.
}
\index{RSA} \label{rsaverfahren}

@@

   \bibitem[Balcazar1988]{Balcazar1988} \index{Balcazar 1988} 
       Balcazar J.L., Daaz J., Gabarr\'{} J.: \\ 
       \emph{Structural Complexity I.} \\ 
       Springer Verlag, pp 63.

    \bibitem[Brickell1985]{Brickell1985} \index{Brickell 1985}
       Brickell E.F.: \\
       \emph{Breaking Iterated Knapsacks.}  \\
       Advances in Cryptology: Proc. CRYPTO\'84, 
       Lecture Notes in Computer Science, vol. 196, 
       Springer-Verlag, New York, 1985, pp. 342-358.

>>>>

%--------------------------------------------------------------------
\hypertarget{RSAVerfahren}{}
\subsubsection[The RSA procedure]
{The RSA procedure\footnote{%
Please compare chapters \ref{rsabeweis}, ff.
}$^,$\footnote{%
Using CrypTool\index{CrypTool} you can gain practical experience
with the RSA procedure via the menu {\bf Indiv.Procedures \textbackslash{} 
RSA Cryptosystem \textbackslash{} RSA Demonstration}.
}}

% LateX-Learnings: Mit \footnotemark statt \footnote
% stand der höhere Index unten zweimal (oben ok).
%
%  {The RSA procedure\footnotemark$^,$\footnotemark}
%  \footnotetext{
%  Please compare chapters \ref{rsabeweis}, ff.
%  }
%  \footnotetext{
%  Using CrypTool\index{CrypTool} you can gain practical experience
%  with the RSA procedure via the menu {\bf Indiv.Procedures \textbackslash{} 
%  RSA Cryptosystem \textbackslash{} RSA Demonstration}.
%  }
%  
%   in the article ``Introduction to Elementary Number Theory with Examples'',
%   chapter~\ref{patternsandstructures}
%   \hyperlink{RSABeweis}{Beweis des RSA-Verfahrens mit Euler-Fermat}

\index{RSA} \label{rsaverfahren}

@@

   \bibitem[Balcazar1988]{Balcazar1988} \index{Balcazar 1988} 
       Balcazar J.L., Daaz J., Gabarr J.: \\ 
       \emph{Structural Complexity I.} \\ 
       Springer Verlag, pp 63.

    \bibitem[Brickell1985]{Brickell1985} \index{Brickell 1985}
       Brickell E.F.: \\
       \emph{Breaking Iterated Knapsacks.}  \\
       Advances in Cryptology: Proc. CRYPTO'84, 
       Lecture Notes in Computer Science, vol. 196, 
       Springer-Verlag, New York, 1985, pp. 342-358.
